---
output:
  html_document: default
  pdf_document: default
---
# Adding a directory and datafile

```{r}
getwd()

```
```{r}
seasnail <- read.csv("~/praccomp2024/Praccomp2024/tidydata_SeaSnail_2024.csv")
#seasnail
```
## Checking to make sure each column variable is the correct type of data format.
```{r}
typeof(seasnail$site_id)
#seasnail$site_id
typeof(seasnail$date)
#seasnail$date
typeof(seasnail$visit_no)
#seansnail$visit_no
typeof(seasnail$season)
typeof(seasnail$sex)
# 0 = male, 1 = female
typeof(seasnail$shell_height_mm)
typeof(seasnail$infection_status)
# 0 = uninfected, 1 = infected
typeof(seasnail$trematode_species)
typeof(seasnail$snail_quadrat_no)
typeof(seasnail$snail_density)
typeof(seasnail$snail_density)
typeof(seasnail$lt_height_ft)
typeof(seasnail$water_temp_C)
typeof(seasnail$air_temp_C)
typeof(seasnail$salinity_ppt)
typeof(seasnail$water_depth_cm)
typeof(seasnail$quadrat_distance_m)
typeof(seasnail$shoot_density)
typeof(seasnail$grass_species_1)
typeof(seasnail$no_blades_1)
typeof(seasnail$grass_species_2)
typeof(seasnail$no_blades_2)
```

```{r}
#seasnail
```
## Adding tidyverse
```{r}
str(seasnail)
library(tidyverse)
```
## Removal of all fully empty columns
```{r}
#seasnail %>% dplyr::select(-blade1_length_cm_1, -blade2_length_cm_1, -blade3_length_cm_1, -blade4_length_cm_1, -blade5_length_cm_1, -blade1_length_cm_2, -blade2_length_cm_2, -blade3_length_cm_2, -blade4_length_cm_2, -blade5_length_cm_2, -no_blades_1, -no_blades_2, -tin_wt_g, -dry_filter_wt_g, -tin_dry_filter_wt_g, -wet_filter_wt_g, -tin_filter_epibiont_wt_g, -epibiont_biomass_mg, -foil_wt_g, -grass_quadrat_no)
```

## Filtering 

### Filtering for comparing salinity measurements across all sites in all sample seasons: early summer, late summer, and fall
```{r}
seasnail_season_salinity <- seasnail %>% dplyr::select(site_id, season, date, salinity_ppt) %>%
  filter(!is.na(salinity_ppt)) %>%
  mutate(date = ymd(date))
seasnail_season_salinity
```
### Plot a general linear model for salinity per season by date
```{r}
seasnail_season_salinity_bars <- ggplot(seasnail_season_salinity, aes(x = date, y = salinity_ppt, fill = season)) + 
         labs(x = "Date", y = "Salinity (ppt)", title = "Seasonal Salinity over Time") +
  geom_col(aes(color = season)) +
  theme_linedraw() +
  facet_wrap( ~ site_id)
seasnail_season_salinity_bars
```

```{r}
ggsave("data/seasnail_season_salinity_bars.png")
```


```{r}
library(ggplot2)
#install.packages("ggExtra")
library(ggExtra)
#install.packages("paletteer")
library(paletteer)
#install.packages("ggstatsplot")
library(ggstatsplot)
#install.packages("statsExpressions")
library(statsExpressions)

ggplot(seasnail_season_salinity, aes(x = date, y = salinity_ppt, color = site_id, size = season)) +
      geom_point() +
  theme_ggstatsplot () +
  scale_colour_paletteer_d("MetBrewer::Hiroshige") +
  labs(x = "Date", y = "Salinity (ppt)", title = "Seasonal Salinity Per Site over Time", size = "Season", color = "Site ID") +
  theme(legend.position="right") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(limits = c(24,34))

```

```{r}
ggsave("data/seasnail_season_salinity.png")
```


```{r}
### Filtering and plotting for comparing water temperature measurements across all sites in all sample seasons: early summer, late summer, and fall
```

```{r}
seasnail_season_watertemp <- seasnail %>% dplyr::select(site_id, season, date, water_temp_C) %>%
  filter(!is.na(water_temp_C)) %>%
  mutate(date = ymd(date))
seasnail_season_watertemp
```
### Plot a general linear model for water temperature per season by date
```{r}
ggplot(seasnail_season_watertemp, aes(x = date, y = water_temp_C, fill = season)) +
  labs(x = "Date", y = "Water Temperature (C)", title = "Seasonal Water Temperature over Time") +
  geom_col(aes(color = season)) +
  theme_linedraw() +
  facet_wrap( ~ site_id)
```
```{r}
ggplot(seasnail_season_watertemp, aes(x = date, y = water_temp_C, color = site_id, size = season)) +
      geom_point() +
  theme_ggstatsplot() +
  scale_colour_paletteer_d("MetBrewer::Hiroshige") +
  labs(x = "Date", y = "Water Temperature (C)", title = "Seasonal Water Temperature Per Site over Time", size = "Season", color = "Site ID") +
  theme(legend.position="right") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(limits = c(15,35))
```
```{r}
ggsave("data/seasnail_season_watertemp.png")
```

### Filtering and plotting for comparing air temperature measurements across all sites in all sample seasons: early summer, late summer, and fall
```{r}
seasnail_season_airtemp <- seasnail %>% dplyr::select(site_id, season, date, air_temp_C) %>%
  filter(!is.na(air_temp_C)) %>%
  mutate(date = ymd(date))
seasnail_season_airtemp
```
### Plot a general linear model for water temperature per season by date
```{r}
ggplot(seasnail_season_airtemp, aes(x = date, y = air_temp_C, fill = season)) +
  labs(x = "Date", y = "Air Temperature (C)", title = "Seasonal Air Temperature over Time") +
  geom_col(aes(color = season)) +
  theme_linedraw()+
  facet_wrap( ~ site_id)
```
```{r}
ggplot(seasnail_season_airtemp, aes(x = date, y = air_temp_C, color = site_id, size = season)) +
      geom_point() +
  theme_ggstatsplot() +
  scale_colour_paletteer_d("MetBrewer::Hiroshige") +
  labs(x = "Date", y = "Air Temperature (C)", title = "Seasonal Air Temperature Per Site over Time", size = "Season", color = "Site ID") +
  theme(legend.position="right") +
  scale_x_date(date_breaks = "1 month", date_labels = "%b") +
  scale_y_continuous(limits = c(15,35))
```

```{r}
ggsave("data/seasnail_season_airtemp.png")
```


### Filtering and plotting for comparing low tide height across all sites in all sample seasons: early summer, late summer, and fall
```{r}
seasnail_season_ltheight <- seasnail %>% dplyr::select(site_id, season, date, lt_height_ft) %>%
  filter(!is.na(lt_height_ft)) %>%
  mutate(date = ymd(date))
seasnail_season_ltheight
```
### Plot a general linear model for low tide height per season by date
```{r}
ggplot(seasnail_season_ltheight, aes(x = date, y = lt_height_ft, fill = season)) +
  labs(x = "Date", y = "Low Tide Height (ft)", title = "Seasonal Low Tide Heights over Time") +
  geom_col(aes(color = season)) +
  theme_linedraw()+
  facet_wrap( ~ site_id)
```
### Filtering and plotting for comparing water depth across all sites in all sample seasons: early summer, late summer, and fall
```{r}
seasnail_season_waterdepth <- seasnail %>% dplyr::select(site_id, season, date, water_depth_cm) %>%
  filter(!is.na(water_depth_cm)) %>%
  mutate(date = ymd(date))
seasnail_season_waterdepth
```
### Plot a general linear model for water depth per season by date
```{r}
ggplot(seasnail_season_waterdepth, aes(x = date, y = water_depth_cm, fill = season)) +
  labs(x = "Date", y = "Water Depth (mm)", title = "Seasonal Water Depth at Low Tide per Site over Time") +
  geom_point(aes(color = season)) +
  geom_smooth(method = "lm", color="black", size=0.5) +
  theme_linedraw()
```
###Linear Models

```{r}
getwd()
```
```{r}
setwd("/Users/meghannadzam/Desktop/Datasets and Figs")
```


```{r}
(seasnail_lm <- read.csv("seasnail_2024_lm_prevalence_temp_salinity.csv"))
```

```{r}
seasnail_lm_prev_sal <- seasnail_lm %>% dplyr::select(site_id, date, infection_prevalence, salinity_ppt) %>%
  filter(!is.na(salinity_ppt)) %>%
  mutate(date = ymd(date))
seasnail_lm_prev_sal
```

```{r}
library(ggplot2)
ggplot(seasnail_lm_prev_sal, aes(x = salinity_ppt, y = infection_prevalence)) +
  geom_smooth(method='lm', se=TRUE, color='turquoise4', fill = "lightgrey") +
  geom_point(aes(color = site_id), size=3) +
  theme_ggstatsplot() +
  labs(x='Salinity (ppt)', y='Infection Prevalence (%)', title='Linear Regression of 2024 Water Temperature and Trematode Prevalence', color = "Site ID") +
  theme(plot.title = element_text(hjust=0.5, size=13, face='bold')) +
  theme(legend.position="right") +
  scale_colour_paletteer_d("MetBrewer::Hiroshige") +
  scale_y_continuous(limits = c(0,25))

summary(seasnail_lm_prev_sal)
model <- lm(infection_prevalence ~ salinity_ppt, data=seasnail_lm_prev_sal)
summary(model)

```
```{r}
ggsave("Datasets and Figs/seasnail_lm_prev_sal.png")
```

```{r}
seasnail_lm_prev_temp <- seasnail_lm %>% dplyr::select(site_id, date, infection_prevalence, water_temp_C) %>%
  filter(!is.na(water_temp_C)) %>%
  mutate(date = ymd(date))
seasnail_lm_prev_temp
```
```{r}
ggplot(seasnail_lm_prev_temp, aes(x =water_temp_C, y = infection_prevalence)) +
  geom_smooth(method='lm', se=TRUE, color='turquoise4', fill = "lightgrey") +
  geom_point(aes(color = site_id), size=3) +
  theme_ggstatsplot() +
  labs(x='Water Temperature (C)', y='Infection Prevalence (%)', title='Linear Regression of 2024 Salinity and Trematode Prevalence', color = "Site ID") +
  theme(plot.title = element_text(hjust=0.5, size=13, face='bold')) +
  theme(legend.position="right") +
  scale_colour_paletteer_d("MetBrewer::Hiroshige") + 
  scale_y_continuous(limits = c(0,25))

summary(seasnail_lm_prev_temp)
model <- lm(infection_prevalence ~ water_temp_C, data=seasnail_lm_prev_temp)
summary(model)
```

```{r}
ggsave("Datasets and Figs/seasnail_lm_prev_temp.png")
```
```{r}
citation()

version$version.string

```
```{r}
citation("ggplot2")
packageVersion("ggplot2")
```
```{r}
devtools::session_info()
```
















