# Adding a directory and datafile

```{r}
getwd()

```
```{r}
seasnail <- read.csv("~/praccomp2024/Praccomp2024/tidydata_SeaSnail_2024.csv")
seasnail
```
## Checking to make sure each column variable is the correct type of data format.
```{r}
typeof(seasnail$site_id)
#seasnail$site_id
typeof(seasnail$date)
#seasnail$date
typeof(seasnail$visit_no)
#seansnail$visit_no
typeof(seasnail$season)
typeof(seasnail$sex)
# 0 = male, 1 = female
typeof(seasnail$shell_height_mm)
typeof(seasnail$infection_status)
# 0 = uninfected, 1 = infected
typeof(seasnail$trematode_species)
typeof(seasnail$snail_quadrat_no)
typeof(seasnail$snail_density)
typeof(seasnail$snail_density)
typeof(seasnail$lt_height_ft)
typeof(seasnail$water_temp_C)
typeof(seasnail$air_temp_C)
typeof(seasnail$salinity_ppt)
typeof(seasnail$water_depth_cm)
typeof(seasnail$quadrat_distance_m)
typeof(seasnail$shoot_density)
typeof(seasnail$grass_species_1)
typeof(seasnail$no_blades_1)
typeof(seasnail$grass_species_2)
typeof(seasnail$no_blades_2)
```

```{r}
seasnail
```
## Adding tidyverse
```{r}
str(seasnail)
library(tidyverse)
```
## Removal of all fully empty columns
```{r}
seasnail %>% dplyr::select(-blade1_length_cm_1, -blade2_length_cm_1, -blade3_length_cm_1, -blade4_length_cm_1, -blade5_length_cm_1, -blade1_length_cm_2, -blade2_length_cm_2, -blade3_length_cm_2, -blade4_length_cm_2, -blade5_length_cm_2, -no_blades_1, -no_blades_2, -tin_wt_g, -dry_filter_wt_g, -tin_dry_filter_wt_g, -wet_filter_wt_g, -tin_filter_epibiont_wt_g, -epibiont_biomass_mg, -foil_wt_g, -grass_quadrat_no)
```

## Filtering 

### Filtering for comparing salinity measurements across all sites in all sample seasons: early summer, late summer, and fall
```{r}
seasnail_season_salinity <- seasnail %>% dplyr::select(site_id, season, date, salinity_ppt) %>%
  filter(!is.na(salinity_ppt)) %>%
  mutate(date = ymd(date))
seasnail_season_salinity
```
### Plot a general linear model for salinity per season by date
```{r}
ggplot(seasnail_season_salinity, aes(x = date, y = salinity_ppt, fill = season)) +
  geom_point(aes(color = season)) +
  geom_smooth(method = "lm")
```

###Filtering and plotting for comparing water temperature measurements across all sites in all sample seasons: early summer, late summer, and fall
```{r}
seasnail_season_watertemp <- seasnail %>% dplyr::select(site_id, season, date, water_temp_C) %>%
  filter(!is.na(water_temp_C)) %>%
  mutate(date = ymd(date))
seasnail_season_watertemp
```
### Plot a general linear model for water temperature per season by date
```{r}
ggplot(seasnail_season_watertemp, aes(x = date, y = water_temp_C, fill = season)) +
  geom_point(aes(color = season)) +
  geom_smooth(method = "lm")
```
###Filtering and plotting for comparing air temperature measurements across all sites in all sample seasons: early summer, late summer, and fall
```{r}
seasnail_season_airtemp <- seasnail %>% dplyr::select(site_id, season, date, air_temp_C) %>%
  filter(!is.na(air_temp_C)) %>%
  mutate(date = ymd(date))
seasnail_season_airtemp
```
### Plot a general linear model for water temperature per season by date
```{r}
ggplot(seasnail_season_airtemp, aes(x = date, y = air_temp_C, fill = season)) +
  geom_point(aes(color = season)) +
  geom_smooth(method = "lm")
```
### Filtering and plotting for comparing low tide height across all sites in all sample seasons: early summer, late summer, and fall
```{r}
seasnail_season_ltheight <- seasnail %>% dplyr::select(site_id, season, date, lt_height_ft) %>%
  filter(!is.na(lt_height_ft)) %>%
  mutate(date = ymd(date))
seasnail_season_ltheight
```
### Plot a general linear model for low tide height per season by date
```{r}
ggplot(seasnail_season_ltheight, aes(x = date, y = lt_height_ft, fill = season)) +
  geom_point(aes(color = season)) +
  geom_smooth(method = "lm")
```
### Filtering and plotting for comparing water depth across all sites in all sample seasons: early summer, late summer, and fall
```{r}
seasnail_season_waterdepth <- seasnail %>% dplyr::select(site_id, season, date, water_depth_cm) %>%
  filter(!is.na(water_depth_cm)) %>%
  mutate(date = ymd(date))
seasnail_season_waterdepth
```
### Plot a general linear model for water depth per season by date
```{r}
ggplot(seasnail_season_waterdepth, aes(x = date, y = water_depth_cm, fill = season)) +
  geom_point(aes(color = season)) +
  geom_smooth(method = "lm")
```







